Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex
updates

We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record names and record field names. In the rest of the
paper, we assume that record fields can be freely reordered:

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \addopt{\rho_i}  \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

During inference we use row-variables $\rho_i$ to represent the flexibility
implied by our interpretation of records in samples. 
When a record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
it may be combined in a list with other samples that contain more fields. The
overall shape inferred must account for the potential that any other fields are not
present in the first sample. The solutions of $\rho_i$ variables is 
discussed in \S\ref{sec:inference}. We assume substitution and row-unification is
defined for these in the usual way , except that as substitutions are made
the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
in the vector. After inference any remaining $\rho_i$ variables are ignored (assumed empty).

@@

@@

What remains to be 
specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
%
\begin{equation*}
\begin{array}{lcl}
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The first few cases of represent primitive values ($i$ for integers, $f$ for floating
point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
data $d_1, \ldots, d_n$ into a shape $\sigma$. We assume each record in each $d_i$ is marked with a fresh row inference
variable $\rho_i$.  We then choose a minimal solution $\theta$ for these row variables
in order to equate the record 
field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
This minimization is implemented via row variable unification, we omit the details here.

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.1em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference}
\label{fig:shape inference}
\end{figure}

@@

\paragraph{Properties.}
The revised algorithm stil finds a shape which is the least upper bound. This means that 
labelled top shape is only inferred when there is no other alternative.

>>>>

We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record names and record field names.

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \rho_i  \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

@@

In the rest of the paper, we assume that record fields can be freely reordered.

During inference we use row-variables $\rho_i$ in record shapes to represent the flexibility
arising from records in samples. For example, when a 
record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
it may be combined through inference with other
samples such as {\small\ttfamily \{ x : 3, y : 4 \}} that contain more fields. The
overall shape inferred must account for the fact that any extra fields are optional,
giving an inferred shape {\small\ttfamily \{ x : int, y : option<int> \}}.
The process of solving $\rho_i$ variables is  discussed in \S\ref{sec:formal-inferval}.

@@

We now specify how we obtain the shape from data. We represent JSON, XML and CSV documents using
the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
%
\begin{equation*}
\begin{array}{lcl}
 d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
\end{array}
\end{equation*}
%
The first few cases of represent primitive values ($i$ for integers, $f$ for floating
point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
list of values in square brackets. A record starts with a name $\nu$, followed by a 
sequence of field assignments $\nu_i \mapsto d_i$.

Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume each record in each $d_i$ is marked with a fresh row inference
variable $\rho_i$.  
We then choose a ground, minimal substitution $\theta$ for row variables.
Because $\rho_i$ variables represent potentially missing fields,
the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
in the vector. 
This must be sufficient to equate the record field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
In practice, $\theta$ is found via row variable unification, we omit the details here.
No $\rho_i$ variables remain after inference as the substitution chosen is ground.

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rclcrclcrcl}
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
\end{array}
\end{equation*}
\noindent
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
 \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
 \\[0.5em]
 \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
 \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
 \\[0.5em]
 \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.1em]
 \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
\end{array}
\end{equation*}
\caption{Shape inference}
\label{fig:shape-inference}
\end{figure}

@@

\paragraph{Properties.}
The revised algorithm still finds a shape which is the least upper bound. This means that 
labelled top shape is only inferred when there is no other alternative.
